Reproduction frequency and offspring survival decline in elephant seals past prime age

### 2024-07-11 paper/paperv3.qmd

# Create POPULATION-LEVEL (i.e. without rand effects) predictions
breed_pred_pop <- ggpredict(breed_age_mod, 
                            terms = c("age10 [all]", "age_cat")) %>% 
  as_tibble() %>% 
  mutate(age = x * 10 + age_senesce,
         age_cat = factor(group, levels = c("Young", "Old"))) %>% 
  filter((age_cat == "Young" & age < age_senesce) |
           (age_cat == "Old" & age >= age_senesce),
         age <= 20) %>% 
  left_join(n_breed, by = "age")

# Create WEIGHTED-AVERAGE predictions using years' random effects
seal_years <- count(sealdat, year_fct)
breed_pred_years <- ggpredict(
  breed_age_mod,
  terms = c("age10 [all]", "age_cat", "year_fct [all]"),
  type = "random",
  interval = "confidence"
) %>% 
  as_tibble() %>% 
  mutate(age = x * 10 + age_senesce,
         age_cat = group,
         year_fct = facet) %>% 
  filter((age_cat == "Young" & age < age_senesce) |
           (age_cat == "Old" & age >= age_senesce),
         age <= 20)
breed_pred_wgt <- breed_pred_years %>% 
  left_join(seal_years, by = "year_fct") %>% 
  group_by(age, age_cat) %>% 
  summarize(predicted = weighted.mean(predicted, n),
            .groups = "drop")

ggplot(breed_pred_pop, aes(age, predicted)) +
  geom_line(aes(group = interaction(age_cat, year_fct)),
            breed_pred_years,
            alpha = 0.1) +
  geom_ribbon(aes(fill = age_cat, ymin = conf.low, ymax = conf.high), 
              alpha = 0.5) +
  geom_line(aes(color = age_cat), linewidth = 1.2) +
  geom_line(aes(color = age_cat),
            breed_pred_wgt, 
            linewidth = 1.2,
            linetype = "dashed") +
  geom_pointrange(aes(y = perc_breed,
                      ymin = perc_breed - se,
                      ymax = perc_breed + se),
                  breed_nonbreed) +
  geom_vline(xintercept = age_senesce - 0.5, linetype = "dashed") +
  scale_y_continuous("Breeding probability", labels = scales::percent) +
  scale_color_manual(values = c("#7fbc41", "#de77ae")) +
  scale_fill_manual(values = c("#7fbc41", "#de77ae")) +
  labs(x = "Age (years)") +
  theme(legend.position = "none") + 
  geom_text(aes(label = n, y = .55))

@@

# Summarize the raw data
surv_recr_summ <- surv_recr_data %>% 
  group_by(age) %>% 
  summarize(n_pups = n(), 
            n_fpups = sum(pupsex == "F"),
            perc_surv = sum(pup_surv_int) / n_pups, 
            se_surv = (perc_surv * (1 - perc_surv) / n_pups)^0.5,
            perc_recr = sum(pup_recr_int) / n_fpups, 
            se_recr = (perc_recr * (1 - perc_recr) / n_fpups)^0.5) %>% 
  filter(age <= 20)

#Create sample size data frame
n_surv <- surv_recr_data %>% 
  group_by(age) %>% 
  count(age)

# Create pup survival POPULATION-LEVEL (i.e. without rand effects) predictions
surv_pred_pop <- ggpredict(
  pup_surv_mod, 
  terms = c("age10 [all]", "age_cat"),
  type = "fixed",
  interval = "confidence"
) %>% 
  as_tibble() %>% 
  mutate(age = x * 10 + age_senesce,
         age_cat = group) %>% 
  filter((age_cat == "Young" & age < age_senesce) |
           (age_cat == "Old" & age >= age_senesce),
         age <= 20) %>% 
  left_join(n_surv, by = "age")

@@

fig_surv <- ggplot(surv_pred_pop, aes(age, predicted)) +
  geom_line(aes(group = interaction(age_cat, year_fct)),
            surv_pred_years,
            alpha = 0.1) +
  geom_ribbon(aes(fill = age_cat, ymin = conf.low, ymax = conf.high), 
              alpha = 0.2) +
  geom_line(aes(color = age_cat, linetype = age_cat), linewidth = 1.2) +
  geom_pointrange(aes(y = perc_surv, 
                      ymin = perc_surv - se_surv, 
                      ymax = perc_surv + se_surv),
                  surv_recr_summ) +
  geom_vline(xintercept = age_senesce - 0.5, linetype = "dashed") +
  scale_y_continuous("Offspring survival", 
                     labels = scales::percent,
                     limits = c(0, 0.68)) +
  scale_color_manual(values = c("#7fbc41", "#de77ae")) +
  scale_fill_manual(values = c("#7fbc41", "#de77ae")) +
  scale_linetype_manual(values = c("dashed", "solid")) +
  labs(x = "Age (years)") +
  theme(legend.position = "none") + 
  geom_text(aes(label = n, y = .65))

@@

# Create pup recruit POPULATION-LEVEL (i.e. without rand effects) predictions
recr_pred_pop <- ggpredict(
  pup_recr_mod, 
  terms = c("age10 [all]", "age_cat"),
  type = "fixed",
  interval = "confidence"
) %>% 
  as_tibble() %>% 
  mutate(age = x * 10 + age_senesce,
         age_cat = group) %>% 
  filter((age_cat == "Young" & age < age_senesce) |
           (age_cat == "Old" & age >= age_senesce),
         age <= 20) %>% 
  left_join(n_recruit, by = "age")

@@

fig_recr <- ggplot(recr_pred_pop, aes(age, predicted)) +
  geom_line(aes(group = interaction(age_cat, year_fct)),
            recr_pred_years,
            alpha = 0.1) +
  geom_ribbon(aes(fill = age_cat, ymin = conf.low, ymax = conf.high), 
              alpha = 0.2) +
  geom_line(aes(color = age_cat), linewidth = 1.2, linetype = "dashed") +
  geom_pointrange(aes(y = perc_recr, 
                      ymin = perc_recr - se_recr, 
                      ymax = perc_recr + se_recr),
                  surv_recr_summ) +
  geom_vline(xintercept = age_senesce - 0.5, linetype = "dashed") +
  scale_y_continuous("Offspring recruitment", 
                     labels = scales::percent,
                     limits = c(0, 0.68)) +
  scale_color_manual(values = c("#7fbc41", "#de77ae")) +
  scale_fill_manual(values = c("#7fbc41", "#de77ae")) +
  labs(x = "Age (years)") +
  theme(legend.position = "none") + 
  geom_text(aes(label = n, y = .65))

@@

# Create POPULATION-LEVEL (i.e. without rand effects) predictions
sexrat_pred <- ggpredict(sexrat_age_mod, 
                         terms = c("age10 [all]", "age_cat")) %>% 
  as_tibble() %>% 
  mutate(age = x * 10 + age_senesce,
         age_cat = factor(group, levels = c("Young", "Old"))) %>% 
  filter((age_cat == "Young" & age < age_senesce) |
           (age_cat == "Old" & age >= age_senesce)) %>% 
  left_join(n_sexrat, by = "age")

# Predictions for each year (random effect)
sexrat_pred_years <- ggpredict(
  sexrat_age_mod,
  terms = c("age10 [all]", "age_cat", "year_fct [all]"),
  type = "random",
  interval = "confidence"
) %>% 
  as_tibble() %>% 
  mutate(age = x * 10 + age_senesce,
         age_cat = group,
         year_fct = facet) %>% 
  filter((age_cat == "Young" & age < age_senesce) |
           (age_cat == "Old" & age >= age_senesce),
         age <= 20)

ggplot(sexrat_pred, aes(age, predicted)) +
  geom_line(aes(group = interaction(age_cat, year_fct)),
            sexrat_pred_years,
            alpha = 0.1) +
  geom_ribbon(aes(fill = age_cat, ymin = conf.low, ymax = conf.high), 
              alpha = 0.2) +
  geom_line(aes(color = age_cat), linewidth = 1.2, linetype = "dashed") +
  geom_pointrange(aes(y = perc_male, 
                      ymin = perc_male - se, 
                      ymax = perc_male + se),
                  sexrat_summ) +
  geom_vline(xintercept = age_senesce - 0.5, linetype = "dashed") +
  scale_color_manual(values = c("#7fbc41", "#de77ae")) +
  scale_fill_manual(values = c("#7fbc41", "#de77ae")) +
  scale_y_continuous("Male offspring", labels = scales::percent) +
  labs(x = "Age (years)") +
  theme(legend.position = "none") +
  geom_text(aes(label = n, y = .95))

>>>>

# # Create POPULATION-LEVEL (i.e. without rand effects) predictions
# breed_pred_pop <- ggpredict(breed_age_mod, 
#                             terms = c("age10 [all]", "age_cat")) %>% 
#   as_tibble() %>% 
#   mutate(age = x * 10 + age_senesce,
#          age_cat = factor(group, levels = c("Young", "Old"))) %>% 
#   filter((age_cat == "Young" & age < age_senesce) |
#            (age_cat == "Old" & age >= age_senesce),
#          age <= 20) %>% 
#   left_join(n_breed, by = "age")

# Create ANNUAL (i.e. rand effect of year only) predictions
breed_pred_years <- ggpredict(
  breed_age_mod, 
  terms = c("age10 [all]", "age_cat", "year_fct"),
  type = "random"
) %>% 
  as_tibble() %>% 
  mutate(age = x * 10 + age_senesce,
         age_cat = factor(group, levels = c("Young", "Old")),
         year_fct = factor(facet)) %>% 
  filter((age_cat == "Young" & age < age_senesce) |
           (age_cat == "Old" & age >= age_senesce),
         age <= 20)

# Create WEIGHTED-AVERAGE predictions using years' random effects
# ggpredict can't do that for confidence intervals
# create prediction grid
breed_grid <- expand_grid(
  age = 4:20,
  year_fct = factor(1988:2023)
) %>% 
  mutate(age10 = (age - age_senesce) / 10,
         age_cat = factor(ifelse(age < age_senesce, "Young", "Old"),
                          levels = c("Young", "Old")))

seal_years <- count(sealdat, year_fct)
breed_pred_wgt <- breed_grid %>% 
  mutate(predicted = predict(breed_age_mod, 
                             newdata = breed_grid,
                             re.form = ~ (1 | year_fct)),
         predicted_se = suppressWarnings(
           # warning: "se.fit computation uses an approximation to estimate the sampling distribution of the parameters"
           predict(breed_age_mod, 
                   newdata = breed_grid,
                   re.form = NA,
                   se.fit = TRUE)
         )$se.fit,
         predicted_pop = predict(breed_age_mod, 
                                 newdata = breed_grid,
                                 re.form = NA),
         conf_lo = predicted - 1.96 * predicted_se,
         conf_hi = predicted + 1.96 * predicted_se) %>% 
  left_join(seal_years, by = "year_fct") %>% 
  group_by(age, age_cat) %>% 
  summarize(across(c(predicted, predicted_pop, conf_lo, conf_hi), 
                   \(x) weighted.mean(x, n)),
            .groups = "drop") %>% 
  # invert logit
  mutate(across(c(predicted, predicted_pop, conf_lo, conf_hi), 
                family(breed_age_mod)$linkinv)) %>% 
    left_join(n_breed, by = "age")

ggplot(breed_pred_wgt, aes(age, predicted)) +
  # Predictions for individual years
  geom_line(aes(group = interaction(age_cat, year_fct)),
            breed_pred_years,
            alpha = 0.1) +
  # Weighted model (CI and mean response)
  geom_ribbon(aes(fill = age_cat, ymin = conf_lo, ymax = conf_hi), 
              alpha = 0.5) +
  geom_line(aes(color = age_cat),
            linewidth = 1.2) +
  # Population-level mean response
  geom_line(aes(y = predicted_pop, group = age_cat),
            linetype = "dotted",
            linewidth = 1.2,
            alpha = 0.5) +
  geom_pointrange(aes(y = perc_breed,
                      ymin = perc_breed - 1.96 * se,
                      ymax = perc_breed + 1.96 * se),
                  breed_nonbreed) +
  geom_vline(xintercept = age_senesce - 0.5, linetype = "dashed") +
  scale_y_continuous("Breeding probability", labels = scales::percent) +
  scale_color_manual(values = c("#7fbc41", "#de77ae")) +
  scale_fill_manual(values = c("#7fbc41", "#de77ae")) +
  labs(x = "Age (years)") +
  theme(legend.position = "none") + 
  geom_text(aes(label = n, y = 1.1))

@@

# Summarize the raw data
surv_recr_summ <- surv_recr_data %>% 
  group_by(age) %>% 
  filter(age <= 19) %>% 
  summarize(n_pups = n(), 
            n_fpups = sum(pupsex == "F"),
            perc_surv = sum(pup_surv_int) / n_pups, 
            se_surv = (perc_surv * (1 - perc_surv) / n_pups)^0.5,
            perc_recr = sum(pup_recr_int) / n_fpups, 
            se_recr = (perc_recr * (1 - perc_recr) / n_fpups)^0.5)

#Create sample size data frame
n_surv <- surv_recr_data %>% 
  group_by(age) %>% 
  filter(age <= 19) %>% 
  count(age)

# Create pup survival predictions (population-level and weighted by year)
surv_grid <- expand_grid(
  age = 4:19,
  year_fct = factor(1988:2016)
) %>% 
  mutate(age10 = (age - age_senesce) / 10,
         age_cat = factor(ifelse(age < age_senesce, "Young", "Old"),
                          levels = c("Young", "Old")))

surv_years <- count(surv_recr_data, year_fct)
surv_pred_wgt <- surv_grid %>% 
  mutate(predicted = predict(pup_surv_mod, 
                             newdata = surv_grid,
                             re.form = ~ (1 | year_fct)),
         predicted_se = suppressWarnings(
           # warning: "se.fit computation uses an approximation to estimate the sampling distribution of the parameters"
           predict(pup_surv_mod, 
                   newdata = surv_grid,
                   re.form = NA,
                   se.fit = TRUE),
         )$se.fit,
         predicted_pop = predict(pup_surv_mod, 
                                 newdata = surv_grid,
                                 re.form = NA),
         conf_lo = predicted - 1.96 * predicted_se,
         conf_hi = predicted + 1.96 * predicted_se) %>% 
  left_join(surv_years, by = "year_fct") %>% 
  group_by(age, age_cat) %>% 
  summarize(across(c(predicted, predicted_pop, conf_lo, conf_hi), 
                   \(x) weighted.mean(x, n)),
            .groups = "drop") %>% 
  # invert logit
  mutate(across(c(predicted, predicted_pop, conf_lo, conf_hi), 
                family(pup_surv_mod)$linkinv)) %>% 
      left_join(n_surv, by = "age")

@@

fig_surv <- ggplot(surv_pred_wgt, aes(age, predicted)) +
  # Individual years
  geom_line(aes(group = interaction(age_cat, year_fct)),
            surv_pred_years,
            alpha = 0.1) +
  # Weighted model (confidence interval and mean response)
  geom_ribbon(aes(fill = age_cat, ymin = conf_lo, ymax = conf_hi), 
              alpha = 0.2) +
  geom_line(aes(color = age_cat, linetype = age_cat), linewidth = 1.2) +
  # Population-level mean response
  geom_line(aes(y = predicted_pop, group = age_cat),
            linetype = "dotted", linewidth = 1.2, alpha = 0.5) +
  # Raw data
  geom_pointrange(aes(y = perc_surv, 
                      ymin = perc_surv - se_surv, 
                      ymax = perc_surv + se_surv),
                  surv_recr_summ) +
  # Senescence threshold
  geom_vline(xintercept = age_senesce - 0.5, linetype = "dashed") +
  scale_y_continuous("Offspring survival", 
                     labels = scales::percent,
                     limits = c(0, 0.68)) +
  scale_color_manual(values = c("#7fbc41", "#de77ae")) +
  scale_fill_manual(values = c("#7fbc41", "#de77ae")) +
  scale_linetype_manual(values = c("dashed", "solid")) +
  labs(x = "Age (years)") +
  theme(legend.position = "none") + 
  geom_text(aes(label = n, y = .65))

@@

# Create pup recruitment predictions (population-level and weighted by year)
recr_grid <- expand_grid(
  age = 4:19,
  year_fct = factor(1988:2016)
) %>% 
  mutate(age10 = (age - age_senesce) / 10,
         age_cat = factor(ifelse(age < age_senesce, "Young", "Old"),
                          levels = c("Young", "Old")))

recr_years <- surv_recr_data %>% 
  filter(pupsex == "F") %>% 
  count(year_fct)
recr_pred <- recr_grid %>% 
  mutate(predicted = predict(pup_recr_mod, 
                             newdata = recr_grid,
                             re.form = ~ (1 | year_fct)),
         predicted_se = suppressWarnings(
           # warning: "se.fit computation uses an approximation to estimate the sampling distribution of the parameters"
           predict(pup_recr_mod, 
                   newdata = recr_grid,
                   re.form = NA,
                   se.fit = TRUE),
         )$se.fit,
         predicted_pop = predict(pup_recr_mod, 
                                 newdata = recr_grid,
                                 re.form = NA),
         conf_lo = predicted - 1.96 * predicted_se,
         conf_hi = predicted + 1.96 * predicted_se) %>% 
  left_join(recr_years, by = "year_fct") %>% 
  group_by(age, age_cat) %>% 
  summarize(across(c(predicted, predicted_pop, conf_lo, conf_hi), 
                   \(x) weighted.mean(x, n)),
            .groups = "drop") %>% 
  # invert logit
  mutate(across(c(predicted, predicted_pop, conf_lo, conf_hi), 
                family(pup_recr_mod)$linkinv)) %>% 
        left_join(n_recruit, by = "age")

@@

fig_recr <- ggplot(recr_pred, aes(age, predicted)) +
  # Individual years
  geom_line(aes(group = interaction(age_cat, year_fct)),
            recr_pred_years,
            alpha = 0.1) +
  # Weighted model (confidence interval and mean response)
  geom_ribbon(aes(fill = age_cat, ymin = conf_lo, ymax = conf_hi), 
              alpha = 0.2) +
  geom_line(aes(color = age_cat), linetype = "dashed", linewidth = 1.2) +
  # Population-level mean response
  geom_line(aes(y = predicted_pop, group = age_cat),
            linetype = "dotted", linewidth = 1.2, alpha = 0.5) +
  # Raw data
  geom_pointrange(aes(y = perc_recr, 
                      ymin = perc_recr - se_recr, 
                      ymax = perc_recr + se_recr),
                  surv_recr_summ) +
  geom_vline(xintercept = age_senesce - 0.5, linetype = "dashed") +
  scale_y_continuous("Offspring recruitment", 
                     labels = scales::percent,
                     limits = c(0, 0.68)) +
  scale_color_manual(values = c("#7fbc41", "#de77ae")) +
  scale_fill_manual(values = c("#7fbc41", "#de77ae")) +
  labs(x = "Age (years)") +
  theme(legend.position = "none") + 
  geom_text(aes(label = n, y = .65))

@@

# Create offspring sex ratio predictions (population-level and weighted by year)
sexrat_grid <- expand_grid(
  age = 4:18,
  year_fct = factor(1988:2023)
) %>% 
  mutate(age10 = (age - age_senesce) / 10,
         age_cat = factor(ifelse(age < age_senesce, "Young", "Old"),
                          levels = c("Young", "Old")))

sexrat_years <- count(pup_sex_data, year_fct)
sexrat_pred <- sexrat_grid %>% 
  mutate(predicted = predict(sexrat_age_mod, 
                             newdata = sexrat_grid,
                             re.form = ~ (1 | year_fct)),
         predicted_se = suppressWarnings(
           # warning: "se.fit computation uses an approximation to estimate the sampling distribution of the parameters"
           predict(sexrat_age_mod, 
                   newdata = sexrat_grid,
                   re.form = NA,
                   se.fit = TRUE),
         )$se.fit,
         predicted_pop = predict(sexrat_age_mod, 
                                 newdata = sexrat_grid,
                                 re.form = NA),
         conf_lo = predicted - 1.96 * predicted_se,
         conf_hi = predicted + 1.96 * predicted_se) %>% 
  left_join(sexrat_years, by = "year_fct") %>% 
  group_by(age, age_cat) %>% 
  summarize(across(c(predicted, predicted_pop, conf_lo, conf_hi), 
                   \(x) weighted.mean(x, n)),
            .groups = "drop") %>% 
  # invert logit
  mutate(across(c(predicted, predicted_pop, conf_lo, conf_hi), 
                family(sexrat_age_mod)$linkinv)) %>% 
  left_join(n_sexrat, by = "age")

# Predictions for each year (random effect)
sexrat_years <- ggpredict(
  sexrat_age_mod,
  terms = c("age10 [all]", "age_cat", "year_fct [all]"),
  type = "random",
  interval = "confidence"
) %>% 
  as_tibble() %>% 
  mutate(age = x * 10 + age_senesce,
         age_cat = group,
         year_fct = facet) %>% 
  filter((age_cat == "Young" & age < age_senesce) |
           (age_cat == "Old" & age >= age_senesce),
         age <= 20)

ggplot(sexrat_pred, aes(age, predicted)) +
  # Individual years
  geom_line(aes(group = interaction(age_cat, year_fct)),
            sexrat_years,
            alpha = 0.1) +
  # Weighted model (confidence interval and mean response)
  geom_ribbon(aes(fill = age_cat, ymin = conf_lo, ymax = conf_hi), 
              alpha = 0.2) +
  geom_line(aes(color = age_cat, group = age_cat),
            linetype = "dashed", linewidth = 1.2) +
  # Population-level mean response
  geom_line(aes(y = predicted_pop, group = age_cat),
            linetype = "dotted", linewidth = 1.2, alpha = 0.5) +
  # Raw data
  geom_pointrange(aes(y = perc_male, 
                      ymin = perc_male - se, 
                      ymax = perc_male + se),
                  sexrat_summ) +
  geom_vline(xintercept = age_senesce - 0.5, linetype = "dashed") +
  scale_y_continuous("Male offspring", labels = scales::percent) +
  scale_color_manual(values = c("#7fbc41", "#de77ae")) +
  scale_fill_manual(values = c("#7fbc41", "#de77ae")) +
  labs(x = "Age (years)") +
  theme(legend.position = "none") +
  geom_text(aes(label = n, y = .95))
